---
title: "proteomicAge: clocks, Global Age, and input formats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{proteomicAge clocks and Global Age}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(proteomicAge)
```

# Overview

`proteomicAge` computes biological age from plasma proteomic data. SomaScan
clock protein columns can be named as gene symbols, UniProt accessions,
SomaScan SL IDs, or SomaScan dot-format SeqIds. Olink-based clocks expect
Olink NPX values with protein columns named by gene symbols.

The package includes five conventional proteomic aging clocks:

| Clock | Function | Main input format |
|---|---|---|
| Tanaka 2018 | `compute_tanaka2018_age()` | UniProt, gene, SL ID, or SeqId |
| Lehallier 2019 | `compute_lehallier2019_age()` | UniProt, gene, SL ID, or SeqId |
| Sathyan 2020 | `compute_sathyan2020_age()` | UniProt, gene, SL ID, or SeqId |
| Oh 2023 conventional | `compute_oh2023_conventional_age()` | SeqId recommended |
| Wang 2024 ARIC midlife | `compute_wang2024_aric_age()` | SeqId recommended |

The package also includes Olink-based models:

| Clock | Function | Main input format |
|---|---|---|
| Kuo 2024 PAC | `compute_kuo2024_pac_age()` | Olink NPX, gene symbols |
| Goeminne organAging | `compute_goeminne2025_organ_age()` | Olink NPX, gene symbols |

Goeminne organAging coefficients are included for academic/non-commercial use
under the source repository license.

It also includes the **Global proteomic age ensemble**:

```r
compute_global_age()
```

`compute_global_age()` implements the Global proteomic clock from Robinson
et al. (2026, Nature Aging). It averages predicted ages and age gaps from the
five conventional clocks.

# Input data

Input data should have one row per sample and one column per protein.

| Column | Required | Notes |
|---|---|---|
| Sample ID | Yes | Set with `id_col` |
| Age | Yes | Chronological age, set with `age_col` |
| Sex | Optional | Set with `sex_col`; `0 = male`, `1 = female` by default |
| Protein columns | Yes | Numeric SomaScan RFU values |

Supported protein naming conventions:

| `match_by` | Example |
|---|---|
| `seqid_dot` | `seq.4374.45`, `4374.45`, `seq.4374.45.2` |
| `seqid_sl` | `SL003869` |
| `gene` | `GDF15` |
| `uniprot` | `P36222` |

Use `detect_format()` when you are not sure how your protein columns are named.

```{r}
detect_format(c("seq.4374.45", "seq.3045.72", "seq.11104.130"))
detect_format(c("GDF15", "PTN", "CHI3L1"))
detect_format(c("SL003340", "SL012561"))
```

# Recommended workflow

For a full SomaScan dataset, the usual workflow is:

```r
dat <- read.csv("my_somascan_data.csv")

protein_cols <- setdiff(names(dat), c("SampleID", "Age", "Sex"))
fmt <- detect_format(protein_cols)

global_age <- compute_global_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  sex_col = "Sex",
  male_value = 0,
  match_by = fmt
)
```

The result contains the Global proteomic age and the component clocks:

| Column | Meaning |
|---|---|
| `proteomic_age` | Mean predicted age across five conventional clocks |
| `age_acceleration` | Mean age gap across five conventional clocks |
| `n_clocks` | Number of component clocks used |
| `*_age` | Component clock predicted age |
| `*_age_acceleration` | Component clock age gap |

# QC summaries and visualizations

The package provides helper functions for QC and comparison of clock outputs.
They can consume either:

- the output from `compute_global_age()`, which already contains Global Age and
  the five component clocks; or
- a manually assembled named list of outputs from the individual
  `compute_*_age()` functions.

For a Global Age result:

```r
qc <- summarize_clock_qc(global_age)
qc

cormat <- clock_correlation_matrix(global_age, value = "age_acceleration")
cormat$correlation
cormat$p_adjusted

plot_clock_correlation_matrix(global_age, value = "age_acceleration")
plot_clock_scatter(global_age, value = "proteomic_age")
plot_clock_scatter(global_age, value = "age_acceleration")
```

For individual clock outputs:

```r
clock_outputs <- list(
  tanaka2018 = tanaka_age,
  lehallier2019 = lehallier_age,
  sathyan2020 = sathyan_age,
  oh2023_conventional = oh_age,
  wang2024_aric = wang_age
)

summarize_clock_qc(clock_outputs)
plot_clock_correlation_matrix(clock_outputs)
```

For datasets with study centers, ancestry groups, ethnicity groups, or
case-control status, `plot_clock_violin()` can stratify clock outputs by a
user-selected categorical variable:

```r
sample_info <- data.frame(
  id = global_age$id,
  center = dat$Center
)

plot_clock_violin(
  global_age,
  group = "center",
  sample_data = sample_info,
  value = "age_acceleration"
)
```

# Running individual clocks

You can also run the clocks separately.

```r
tanaka_age <- compute_tanaka2018_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  match_by = fmt
)

lehallier_age <- compute_lehallier2019_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  sex_col = "Sex",
  male_value = 0,
  match_by = fmt
)

sathyan_age <- compute_sathyan2020_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  sex_col = "Sex",
  match_by = fmt
)

oh_age <- compute_oh2023_conventional_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  sex_col = "Sex",
  match_by = fmt
)

wang_age <- compute_wang2024_aric_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  match_by = fmt
)
```

# Running Olink-based clocks

PAC requires chronological age plus 128 Olink protein predictors:

```r
pac_age <- compute_kuo2024_pac_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  match_by = "gene"
)
```

Goeminne organAging returns one row per sample and organ. Chronological models
return organ-specific predicted ages and age accelerations:

```r
goeminne_age <- compute_goeminne2025_organ_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  organs = c("Conventional", "Heart", "Brain"),
  model_type = "chronological",
  fold = 1
)
```

Mortality-based Goeminne models return `mortality_score`, a relative
log-mortality-hazard score, rather than age in years:

```r
goeminne_mortality <- compute_goeminne2025_organ_age(
  dat,
  id_col = "SampleID",
  age_col = "Age",
  organs = "Conventional",
  model_type = "mortality"
)
```

If an Olink matrix uses UniProt IDs as protein column names, build a map from
the original long Olink table and pass it to the Olink clock functions:

```r
protein_map <- olink_protein_map_from_long(olink_long)

pac_age <- compute_kuo2024_pac_age(
  olink_wide_uniprot,
  id_col = "SampleID",
  age_col = "Age",
  match_by = "uniprot",
  protein_map = protein_map
)

goeminne_heart <- compute_goeminne2025_organ_age(
  olink_wide_uniprot,
  id_col = "SampleID",
  age_col = "Age",
  organs = "Heart",
  match_by = "uniprot",
  protein_map = protein_map
)
```

# Clock protein lists

Each clock has a helper that returns the model proteins.

```{r}
nrow(tanaka2018_proteins())
nrow(lehallier2019_proteins())
nrow(sathyan2020_proteins())
nrow(oh2023_conventional_proteins())
nrow(wang2024_aric_proteins())
nrow(kuo2024_pac_proteins())
length(unique(goeminne2025_organaging_proteins()$protein))
```

# Small executable example

This example uses the Tanaka clock because it has a compact 76-protein model
and can run quickly inside the vignette.

```{r}
set.seed(123)
prots <- tanaka2018_proteins()

demo <- data.frame(
  SampleID = paste0("P", 1:20),
  Age = round(runif(20, 35, 80), 1),
  Sex = sample(c(0, 1), 20, replace = TRUE),
  stringsAsFactors = FALSE
)

for (sid in prots$seqid_sl) {
  demo[[sid]] <- rlnorm(20, meanlog = log(2000), sdlog = 0.5)
}

tanaka_demo <- compute_tanaka2018_age(
  demo,
  id_col = "SampleID",
  age_col = "Age",
  match_by = "seqid_sl"
)

head(tanaka_demo)
```

# Output interpretation

All clock functions return:

| Column | Meaning |
|---|---|
| `id` | Sample identifier |
| `chronological_age` | Input chronological age |
| `proteomic_age` | Predicted proteomic age |
| `age_acceleration` | Age gap / residual age acceleration |
| `n_proteins_matched` | Number of clock proteins found in the input |
| `n_proteins_missing` | Number of clock proteins not found in the input |
| `match_by` | Protein identifier convention used |

In the current implementation, each individual clock's `age_acceleration` is
computed as the residual from:

```r
lm(proteomic_age ~ chronological_age)
```

Positive age acceleration means the sample is predicted to be proteomically
older than expected for its chronological age in that dataset.

# References

Tanaka T, et al. Plasma proteomic signature of age in healthy humans.
*Aging Cell*. 2018;17(5):e12799.

Lehallier B, et al. Undulating changes in human plasma proteome profiles
across the lifespan. *Nature Medicine*. 2019;25(12):1843-1850.

Sathyan S, et al. Plasma proteomic profile of age, health span, and all-cause
mortality in older adults. *Aging Cell*. 2020;19(11):e13250.

Oh HS, et al. Organ aging signatures in the plasma proteome track health and
disease. *Nature*. 2023;624:164-172.

Wang S, et al. Development, characterization, and replication of proteomic aging
clocks: analysis of 2 population-based cohorts. *PLOS Medicine*.
2024;21(9):e1004464.

Robinson O, et al. Associations of proteomic age clocks with lifestyle risk
factors, incident chronic diseases and mortality in two European cohorts.
*Nature Aging*. 2026. doi:10.1038/s43587-026-01163-6.

```{r}
sessionInfo()
```
