---
title: "proteomicAge: Manual & Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manual & Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
library(proteomicAge)
```

# Overview

`proteomicAge` computes biological age from SOMAscan proteomic data.
Protein columns can use gene symbols, `seq.xxx.xx` SeqIds, `SLxxxxxx` SeqIds,
or UniProt accessions. The package auto-detects the naming convention and maps
it to the clock's internal format via `prolist.csv` (7,596 proteins).

# Installation

```r
remotes::install_github("EnricMartin/proteomicAge")
```

# Input Format

Your data.frame should contain:

| Column | Required | Description |
|--------|----------|-------------|
| Sample ID | Yes | Unique sample identifier (any name) |
| Age | Yes | Chronological age in years |
| Sex | No | 0 = male, 1 = female (for Lehallier clock) |
| Protein columns | Yes | Named using one of the supported conventions |

**Supported protein naming:**

| Convention | Example |
|---|---|
| Gene symbol | `GDF15`, `CHI3L1`, `SOST` |
| SeqId dot | `seq.4374.45.2`, `11104.13.3` |
| SeqId SL | `SL003869`, `SL003340` |
| UniProt | `P36222`, `Q99988` |

# Function Reference

## `tanaka2018_proteins()` / `lehallier2019_proteins()`

```{r}
head(tanaka2018_proteins()[, c("Gene", "UniProt", "Weight")], 5)
```

## `detect_format()`

Auto-detect column naming convention.

```{r}
detect_format(c("GDF15", "CHI3L1"))
detect_format(c("P36222", "Q99988"))
detect_format(c("SL003869"))
detect_format(c("seq.4374.45.2"))
```

## `convert_format()`

Convert between naming conventions.

```{r}
data <- data.frame(SampleID="S1", Age=50, GDF15=5000)
convert_format(data, target_format="uniprot")
```

## `compute_tanaka2018_age()`

76-protein elastic net clock. logŌéé transform.

```{r}
data <- data.frame(
  SampleID = paste0("S", 1:3),
  Age = c(40, 60, 80),
  GDF15  = c(2000, 5000, 12000),
  CHI3L1 = c(1500, 3000, 6000),
  SOST   = c(1000, 2500, 5000)
)
compute_tanaka2018_age(data, match_by = "gene")
```

**Parameters:**

| Parameter | Default | Description |
|---|---|---|
| `data` | (required) | data.frame with protein columns + Age |
| `id_col` | `"SampleID"` | ID column name |
| `age_col` | `"Age"` | Age column name |
| `log_transform` | `TRUE` | Apply logŌéé transform |
| `match_by` | `"uniprot"` | How to match: `"gene"`, `"uniprot"`, `"seqid_sl"`, `"seqid_dot"` |

## `compute_lehallier2019_age()`

373-protein LASSO clock. logŌéüŌéĆ transform.

```{r}
compute_lehallier2019_age(data, match_by = "gene")
```

Additional parameters: `sex_col = "Sex"`, `male_value = 0`.

## `preprocess_somascan()`

```{r}
raw <- data.frame(SampleID=1:5, GDF15=c(800,15000,300,2000,5000))
preprocess_somascan(raw, report_missingness=FALSE)
```

# Complete Pipeline

## Step 1: Load & Detect

```{r}
set.seed(123)
prots <- tanaka2018_proteins()
raw <- data.frame(SampleID=paste0("P",1:30), Age=round(runif(30,25,85),1))
for(i in 1:nrow(prots))
  raw[[prots$Gene[i]]] <- rlnorm(30, log(2000), 0.5)

fmt <- detect_format(setdiff(names(raw), c("SampleID","Age")))
cat("Detected format:", fmt, "\n")
```

## Step 2: Compute

```{r}
tanaka_age <- compute_tanaka2018_age(raw, match_by=fmt)
leh_age <- compute_lehallier2019_age(raw, match_by=fmt)
```

## Step 3: Visualize

```{r, fig.width=10, fig.height=5}
par(mfrow=c(1,2))
plot(tanaka_age$chronological_age, tanaka_age$proteomic_age,
     main="Tanaka 2018", xlab="Chronological", ylab="Predicted", pch=19,
     col=rgb(0.2,0.4,0.6,0.5))
abline(0,1,lty=2,col="red")

hist(tanaka_age$age_acceleration, breaks=20, main="Age Acceleration",
     xlab="PROaccel", col="steelblue", border="white")
abline(v=0,col="red",lwd=2)
```

# Understanding Output

| Column | Description |
|--------|-------------|
| `proteomic_age` | Predicted biological age |
| `age_acceleration` | Residuals of `lm(PROage ~ Age)`. Mean Ōēł 0 |
| `n_proteins_matched` | Clock proteins found |
| `n_proteins_missing` | Clock proteins not found |
| `match_by` | Naming convention used for matching |

# Log Transform

| Clock | Transform |
|-------|-----------|
| Tanaka 2018 | logŌéé |
| Lehallier 2019 | logŌéüŌéĆ |

# Session Info

```{r}
sessionInfo()
```
