---
title: "proteomicAge: Manual & Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manual & Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
library(proteomicAge)
```

# Overview

`proteomicAge` computes biological age from SOMAscan proteomic data.
Protein columns can use gene symbols, `seq.xxx.xx` SeqIds, `SLxxxxxx` SeqIds,
or UniProt accessions. The package auto-detects the naming convention and maps
it to the clock's internal format via `prolist.csv` (7,596 proteins).

**Contributed by:** Han Xiao (hx624@ic.ac.uk), Esther Herrera, Arias Julian,
Juan-Carlos Rivilla, Oliver Robinson.

# Installation

```r
remotes::install_github("EnricMartin/proteomicAge")
```

# Input Format

| Column | Required | Description |
|--------|----------|-------------|
| Sample ID | Yes | Unique sample identifier |
| Age | Yes | Chronological age in years |
| Sex | No | 0 = male, 1 = female (Lehallier clock) |
| Protein columns | Yes | Named using one of the conventions below |

**Supported protein naming:**

| Convention | Example |
|---|---|
| Gene symbol | `GDF15`, `CHI3L1`, `SOST` |
| SeqId dot | `seq.4374.45.2`, `11104.13.3` |
| SeqId SL | `SL003869`, `SL003340` |
| UniProt | `P36222`, `Q99988` |

# Step-by-Step Usage

## Step 1: Detect Format

```{r}
protein_cols <- c("GDF15", "CHI3L1", "SOST", "PTN", "ADAMTS5",
                  "MMP12", "EGFR", "IGFBP3", "CLU", "ALB")
detect_format(protein_cols)
```

## Step 2: Convert Format (Optional)

```{r}
data <- data.frame(
  SampleID = paste0("P", 1:5),
  Age = c(40, 55, 62, 71, 35),
  Sex = c(0, 1, 0, 1, 0),
  GDF15 = c(2000, 5000, 9000, 15000, 1800),
  CHI3L1 = c(1500, 3000, 5500, 8000, 1200),
  SOST = c(1000, 2500, 4500, 7000, 900),
  PTN = c(1800, 2800, 5000, 8500, 1600),
  ADAMTS5 = c(2200, 4000, 6500, 10000, 2000),
  MMP12 = c(800, 2000, 3500, 5500, 700),
  EGFR = c(3000, 4500, 4000, 6000, 2800),
  IGFBP3 = c(5000, 3500, 2500, 1800, 5500),
  CLU = c(12000, 15000, 18000, 20000, 10000),
  ALB = c(35000, 38000, 32000, 30000, 40000)
)

# Convert gene symbols to UniProt accessions
data_up <- convert_format(data, target_format = "uniprot",
                          id_col = "SampleID", age_col = "Age")
head(names(data_up))
```

## Step 3: Compute Proteomic Age

```{r}
# Tanaka 2018 — with explicit column name parameters
result_tanaka <- compute_tanaka2018_age(
  data,
  id_col = "SampleID",
  age_col = "Age",
  match_by = "gene"
)
print(result_tanaka[, c("id", "chronological_age", "proteomic_age", "age_acceleration")])
```

```{r}
# Lehallier 2019 — includes sex as covariate
result_leh <- compute_lehallier2019_age(
  data,
  id_col = "SampleID",
  age_col = "Age",
  sex_col = "Sex",
  match_by = "gene"
)
print(result_leh[, c("id", "chronological_age", "proteomic_age", "age_acceleration")])
```

# Function Reference

## `detect_format()`

```{r}
detect_format(c("GDF15", "CHI3L1"))
detect_format(c("P36222", "Q99988"))
detect_format(c("SL003869"))
detect_format(c("seq.4374.45.2"))
```

## `convert_format()`

```{r}
demo <- data.frame(SampleID = "S1", Age = 50, GDF15 = 5000)
convert_format(demo, target_format = "uniprot",
               id_col = "SampleID", age_col = "Age")
```

## `compute_tanaka2018_age()`

76-protein elastic net clock. log₂ transform.

| Parameter | Default | Description |
|---|---|---|
| `data` | (required) | data.frame with protein columns + Age |
| `id_col` | `"SampleID"` | ID column name |
| `age_col` | `"Age"` | Age column name |
| `log_transform` | `TRUE` | Apply log₂ transform |
| `match_by` | `"uniprot"` | `"gene"`, `"uniprot"`, `"seqid_sl"`, `"seqid_dot"` |

## `compute_lehallier2019_age()`

373-protein LASSO clock. log₁₀ transform.

Additional parameters: `sex_col = "Sex"`, `male_value = 0`.

## `preprocess_somascan()`

```{r}
raw <- data.frame(SampleID = 1:5, SL003869 = c(800, 15000, 300, 2000, 5000))
preprocess_somascan(raw, protein_cols = "SL003869", report_missingness = FALSE)
```

# Complete Pipeline

## Step 1: Load & Detect

```{r}
set.seed(123)
prots <- tanaka2018_proteins()
# Use first 20 proteins for demo
demo_prots <- prots[1:20, ]
raw <- data.frame(
  SampleID = paste0("P", 1:30),
  Age = round(runif(30, 25, 85), 1),
  Sex = sample(c(0, 1), 30, replace = TRUE)
)
for (i in 1:nrow(demo_prots)) {
  raw[[demo_prots$Gene[i]]] <- rlnorm(30, log(2000), 0.5)
}

fmt <- detect_format(setdiff(names(raw), c("SampleID", "Age", "Sex")))
cat("Detected format:", fmt, "\n")
```

## Step 2: Compute

```{r}
tanaka_age <- compute_tanaka2018_age(raw, id_col = "SampleID", age_col = "Age",
                                      match_by = fmt)
leh_age <- compute_lehallier2019_age(raw, id_col = "SampleID", age_col = "Age",
                                      sex_col = "Sex", match_by = fmt)
```

## Step 3: Visualize

```{r, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))
plot(tanaka_age$chronological_age, tanaka_age$proteomic_age,
     main = "Tanaka 2018", xlab = "Chronological", ylab = "Predicted",
     pch = 19, col = rgb(0.2, 0.4, 0.6, 0.5))
abline(0, 1, lty = 2, col = "red")
r <- cor(tanaka_age$chronological_age, tanaka_age$proteomic_age, use = "complete")
legend("topleft", legend = paste("r =", round(r, 3)), bty = "n")

hist(tanaka_age$age_acceleration, breaks = 20, main = "Age Acceleration",
     xlab = "PROaccel", col = "steelblue", border = "white")
abline(v = 0, col = "red", lwd = 2)
```

# Understanding Output

| Column | Description |
|--------|-------------|
| `proteomic_age` | Predicted biological age |
| `age_acceleration` | Residuals of `lm(PROage ~ Age)`. Mean ≈ 0 |
| `n_proteins_matched` | Clock proteins found |
| `n_proteins_missing` | Clock proteins not found |
| `match_by` | Naming convention used |

# Log Transform

| Clock | Transform |
|-------|-----------|
| Tanaka 2018 | log₂ |
| Lehallier 2019 | log₁₀ |

```{r}
sessionInfo()
```
