---
title: "proteomicAge: Manual & Pipeline"
author: "ClawBio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manual & Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
library(proteomicAge)
```

# Overview

`proteomicAge` computes biological age from SOMAscan proteomic data.
**Input columns must be named by UniProt accession.** Mapping to the
clock's required SeqId format happens internally via `prolist.csv`.

# Installation

```r
remotes::install_github("EnricMartin/proteomicAge")
```

# Function Reference

## `tanaka2018_proteins()` / `lehallier2019_proteins()`

```{r}
head(tanaka2018_proteins()[, c("UniProt", "Gene", "Weight")], 5)
```

## `compute_tanaka2018_age()`

Tanaka et al. (2018) 76-protein elastic net clock. log₂ transform.

```{r}
data <- data.frame(
  SampleID = "S1",
  Age = 50,
  P36222 = 3000,  # CHI3L1
  Q99988 = 5000,  # GDF15
  Q9BQB4 = 2000,  # SOST
  P21246 = 4000   # PTN
)
compute_tanaka2018_age(data)
```

## `compute_lehallier2019_age()`

Lehallier et al. (2019) 373-protein LASSO clock. log₁₀ transform.

```{r}
compute_lehallier2019_age(data)
```

## `preprocess_somascan()`

Preprocess raw RFU data (log₂, outlier winsorizing).

```{r}
raw <- data.frame(SampleID=1:5, P36222=c(800,15000,300,2000,5000))
preprocess_somascan(raw, report_missingness=FALSE)
```

## `validate_somascan_input()`

```{r}
validate_somascan_input(data)
```

# Complete Pipeline

## Step 1: Load Data

```{r}
set.seed(123)
prots <- tanaka2018_proteins()
raw <- data.frame(SampleID=paste0("P",1:30), Age=round(runif(30,25,85),1))
for(i in 1:nrow(prots))
  if(prots$UniProt[i]!="")
    raw[[prots$UniProt[i]]] <- rlnorm(30, log(2000), 0.5)
head(raw[,1:5])
```

## Step 2: Validate

```{r}
validate_somascan_input(raw)
```

## Step 3: Compute

```{r}
tanaka_age <- compute_tanaka2018_age(raw)
leh_age <- compute_lehallier2019_age(raw)
```

## Step 4: Analyze

```{r, fig.width=10, fig.height=5}
par(mfrow=c(1,2))
plot(tanaka_age$chronological_age, tanaka_age$proteomic_age,
     main="Tanaka 2018", xlab="Chronological", ylab="Predicted", pch=19,
     col=rgb(0.2,0.4,0.6,0.5))
abline(0,1,lty=2,col="red")

hist(tanaka_age$age_acceleration, breaks=20, main="Age Acceleration",
     xlab="PROaccel", col="steelblue", border="white")
abline(v=0,col="red",lwd=2)
```

## Step 5: Compare Clocks

```{r, fig.width=6, fig.height=6}
merged <- merge(tanaka_age[,c("id","proteomic_age")],
                leh_age[,c("id","proteomic_age")], by="id",
                suffixes=c("_tanaka","_leh"))
plot(merged$proteomic_age_tanaka, merged$proteomic_age_leh,
     main="Clock Comparison", xlab="Tanaka", ylab="Lehallier",
     pch=19, col=rgb(0.6,0.2,0.4,0.5))
abline(0,1,lty=2,col="gray50")
r <- cor(merged$proteomic_age_tanaka, merged$proteomic_age_leh, use="complete")
legend("topleft", legend=paste("r =", round(r,3)), bty="n")
```

# Understanding Output

- **`proteomic_age`**: Predicted biological age (PROage)
- **`age_acceleration`**: Residuals of `lm(PROage ~ chronological_age)`. Mean ≈ 0.
  Positive = predicted older than expected; negative = younger.
- **`n_proteins_matched`**: How many of the clock's proteins were found in the data

# Log Transform

| Clock | Transform |
|-------|-----------|
| Tanaka 2018 | log₂ |
| Lehallier 2019 | log₁₀ |

# Session Info

```{r}
sessionInfo()
```
